---
title: "Data exploration"
author: "Ethel Pruss"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(purrr)
library(dplyr)
library(ggplot2)
```

# Compiling BalanceCorpus data

## Import OpenFace data (cam01 only) and aggregate into df 

Compiles data over all available BalanceCorpus trials/dyads

```{r import-openface}
base_dir <- "BalanceCorpus/OpenFace/OF_output"

# cam01 only, drop "preparation" folders
folders <- list.dirs(base_dir, full.names = TRUE, recursive = FALSE)
folders <- folders[str_detect(basename(folders), "_cam01$")]
folders <- folders[!str_detect(basename(folders), "^preparation")]

# folder name looks like: 103_203_12_1_20250113_152455_doughnut_board_clueGiver_cam01
# dyad = first two ids, session_datetime = date_time block, word = everything between
# session_datetime and role, role = clueGiver/guesser, cam = camXX
# (named session_datetime, not "timestamp", so it doesn't clobber OpenFace's own
# per-frame `timestamp` column, which is real elapsed seconds within the recording)
pattern <- "^(\\d+_\\d+)_.*?(\\d{8}_\\d{6})_(.+)_(clueGiver|guesser)_(cam\\d+)$"

df <- map_dfr(folders, function(folder) {
  fname <- basename(folder)
  csv_path <- file.path(folder, paste0(fname, ".csv"))

  if (!file.exists(csv_path)) {
    warning("Missing csv for: ", fname)
    return(NULL)
  }

  m <- str_match(fname, pattern)
  if (any(is.na(m))) {
    warning("Folder name didn't match expected pattern: ", fname)
    return(NULL)
  }

  data <- read.csv(csv_path, stringsAsFactors = FALSE, strip.white = TRUE)

  data$dyad             <- m[2]
  data$session_datetime <- m[3]
  data$word             <- m[4]
  data$role             <- m[5]
  data$cam              <- m[6]
  # condition = "board" or "ground", read off the end of the word label
  data$condition        <- str_extract(m[4], "(board|ground)$")

  data
})

table(df$dyad, df$role)
```

```{r}
write.csv(df,"OF_df.csv")
```

################ Start from here if you already have the OF_df.csv ######

# Calculating delta gaze and expressivity scores

This uses the pre-extracted OpenFace data (facial action units and gaze angle).

## Load df

```{r}
df <- read.csv("OF_df.csv")
```

## Calculate delta scores

### Gaze, brow and squint deltas

```{r gaze-angle-deltas}
# delta = change from previous frame, within each recording (dyad + session_datetime + role)
# AU04 = brow lowerer (frown/tension), AU07 = lid tightener (squint)
# _r = intensity (0-5 continuous), _c = presence (0/1 binary)
df <- df %>%
  arrange(dyad, session_datetime, role, frame) %>%
  group_by(dyad, session_datetime, role) %>%
  mutate(
    delta_gaze_angle_x = gaze_angle_x - lag(gaze_angle_x),
    delta_gaze_angle_y = gaze_angle_y - lag(gaze_angle_y),
    delta_brow_r        = AU04_r - lag(AU04_r),
    delta_brow_c        = AU04_c - lag(AU04_c),
    delta_squint_r      = AU07_r - lag(AU07_r),
    delta_squint_c      = AU07_c - lag(AU07_c)
  ) %>%
  ungroup()
```

### Expressivity delta (mean of all AU_r deltas)

```{r expressivity-delta}
# every AU intensity column (AU01_r, AU02_r, ... AU45_r)
au_r_cols <- grep("^AU[0-9]+_r$", names(df), value = TRUE)

df <- df %>%
  arrange(dyad, session_datetime, role, frame) %>%
  group_by(dyad, session_datetime, role) %>%
  mutate(
    delta_expressivity = {
      au_deltas <- across(all_of(au_r_cols), ~ . - lag(.))
      rowMeans(au_deltas, na.rm = TRUE)
    }
  ) %>%
  ungroup()
```

# Preliminary visualization

## Smooth deltas before downsampling

```{r smooth-deltas}
# raw frame-to-frame deltas are noisy; apply a centered moving-average
# smoother (15 frames ~ 0.5s at 30fps) within each recording before binning
# increase/decrease k for more/less smoothing (k is in frames, ~30 frames/sec)
smooth_ma <- function(x, k = 15) as.numeric(stats::filter(x, rep(1 / k, k), sides = 2))

df <- df %>%
  arrange(dyad, session_datetime, role, frame) %>%
  group_by(dyad, session_datetime, role) %>%
  mutate(
    delta_gaze_angle_x_smooth = smooth_ma(delta_gaze_angle_x),
    delta_gaze_angle_y_smooth = smooth_ma(delta_gaze_angle_y)
  ) %>%
  ungroup()
```

## Downsample to ~10 samples/sec

```{r downsample-10hz}
# raw data is ~30fps; bin into 0.1s windows using OpenFace's native per-frame
# `timestamp` (real elapsed seconds within the recording) and average the
# smoothed deltas within each bin, giving ~10 points/sec per line instead of ~30
df_binned <- df %>%
  mutate(time_bin = floor(timestamp * 10) / 10) %>%
  group_by(dyad, word, condition, role, time_bin) %>%
  summarise(
    delta_gaze_angle_x = mean(delta_gaze_angle_x_smooth, na.rm = TRUE),
    delta_gaze_angle_y = mean(delta_gaze_angle_y_smooth, na.rm = TRUE),
    .groups = "drop"
  )
```

## Line plots per dyad and word

### Plot deltas: guesser vs. clueGiver, one plot per dyad + word

```{r plot-delta-per-word}
plot_delta <- function(data, dyad_val, word_val, metric, ylab) {
  sub <- dplyr::filter(data, dyad == dyad_val, word == word_val)
  # breaks from the real data range (always starts at 0), not the padded panel limits
  x_breaks <- seq(0, ceiling(max(sub$time_bin, na.rm = TRUE)), by = 1)
  ggplot(sub, aes(x = time_bin, y = .data[[metric]], color = role)) +
    geom_line() +
    scale_x_continuous(breaks = x_breaks) +
    labs(title = paste(dyad_val, "-", word_val), x = "Time (s)", y = ylab)
}

# subset: 1 "_board" word and 1 "_ground" word per dyad
combos <- df_binned %>%
  distinct(dyad, word, condition) %>%
  group_by(dyad, condition) %>%
  slice(1) %>%
  ungroup()

for (i in seq_len(nrow(combos))) {
  d <- combos$dyad[i]
  w <- combos$word[i]
  print(plot_delta(df_binned, d, w, "delta_gaze_angle_x", "Δ gaze_angle_x"))
  print(plot_delta(df_binned, d, w, "delta_gaze_angle_y", "Δ gaze_angle_y"))
}
```


### Plot multi-trial timeseries (smoothed, clueGiver vs guesser)

```{r plot-pair-timeseries}
trial_dyad      <- "103_203"
n_trials        <- 5
trial_condition <- "board"  # "board", "ground", or NULL for either

pair_x <- get_aligned_pair_n_trials(df, trial_dyad, "delta_gaze_angle_x", n_trials = n_trials, condition_val = trial_condition)
pair_y <- get_aligned_pair_n_trials(df, trial_dyad, "delta_gaze_angle_y", n_trials = n_trials, condition_val = trial_condition)

# pair$clueGiver/guesser are plain vectors (no frame/time column), so build a
# time axis from sample index and srate; smooth_ma() reused from the earlier
# smoothing chunk (k = frames of moving-average window, default 15 ~ 0.5s)
plot_pair_timeseries <- function(pair, srate, k = 15, ylab = "delta", title_str = "") {
  cg_smooth <- smooth_ma(pair$clueGiver, k = k)
  gs_smooth <- smooth_ma(pair$guesser, k = k)
  n <- length(pair$clueGiver)
  time_s <- (seq_len(n) - 1) / srate

  plot_df <- data.frame(
    time  = rep(time_s, 2),
    value = c(cg_smooth, gs_smooth),
    role  = rep(c("clueGiver", "guesser"), each = n)
  )

  # downsample to ~10 samples/sec: bin into 0.1s windows and average within each
  plot_df <- plot_df %>%
    mutate(time_bin = floor(time * 10) / 10) %>%
    group_by(role, time_bin) %>%
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
    rename(time = time_bin)

  ggplot(plot_df, aes(x = time, y = value, color = role)) +
    geom_line() +
    labs(x = "Time (s)", y = ylab, title = title_str) +
    theme_classic() +
    theme(panel.grid = element_blank(), plot.background = element_rect(fill = "white", color = NA))
}

trials_label <- paste(pair_x$trials, collapse = ", ")

plot_pair_timeseries(pair_x, srate, ylab = "Δ gaze_angle_x (smoothed)",
                      title_str = paste0("On board - unaffiliated, Δ gaze angle x, trials n: (", n_trials, ")"))
plot_pair_timeseries(pair_y, srate, ylab = "Δ gaze_angle_y (smoothed)",
                      title_str = paste0("On board - unaffiliated, Δ gaze angle y, trials n: (", n_trials, ")"))
```

# WMS analysis and plots

```{r}
library(ggplot2)
library(patchwork)
library(wms)
library(remotes)
set.seed(55)
```


## Parameter description for reference 
(This is copied from Travis's slides for reference, the actual parameters used here are defined in the code)
  x, y,                      # two numeric vectors of equal length
  srate          = 100,      # sampling rate (Hz)
  win            = 200,      # window size (samples); >= 100 recommended
  minfreq        = 0.05,     # minimum frequency to resolve (Hz)
  maxfreq        = 50,       # maximum frequency (Hz); <= Nyquist (srate / 2) (half of sampling rate)
  mincycle       = 5,        # min Morlet wavelet cycles
  maxcycle       = 8,        # max Morlet wavelet cycles
  nfreqs         = 40,       # number of frequency bins
  wavelet_length = 4         # wavelet duration (seconds)


## Compiling data in pairs for WMS plots

Note that there is a single trial and multi-trial version. WMS plotting seems more informative over multipel trials, one trial might be too short for cyclical patterns to emerge.

```{r wms-setup}
# native frame rate (frames/sec) from OpenFace's per-frame `timestamp` column
srate <- 30

# for one dyad+word trial: pull clueGiver's and guesser's series for `metric`
# (their two simultaneously-recorded cams), ordered by frame, and trim both
# to the same length so they start and end together
get_aligned_pair <- function(data, dyad_val, word_val, metric) {
  trial <- data %>% filter(dyad == dyad_val, word == word_val)
  cg <- trial %>% filter(role == "clueGiver") %>% arrange(frame) %>% pull(.data[[metric]])
  gs <- trial %>% filter(role == "guesser")   %>% arrange(frame) %>% pull(.data[[metric]])
  # drop each series' first frame: its delta is NA (nothing to diff against),
  # and an NA anywhere breaks the wavelet transform in win_ms_synch/plot_wms
  cg <- cg[-1]
  gs <- gs[-1]
  n  <- min(length(cg), length(gs))
  list(clueGiver = cg[seq_len(n)], guesser = gs[seq_len(n)])
}

# multi-trial version: same idea, but concatenates the first `n_trials` trials
# (words) for a dyad instead of just one. Trials are picked in chronological
# order via `session_datetime` (the trial's real-world start time -- not the
# per-frame `timestamp`, which resets to 0 at the start of every trial and so
# can't be used to order trials against each other). Each trial is aligned and
# trimmed to equal cg/gs length independently (so clueGiver and guesser start
# together within every trial), *then* concatenated in time order -- since
# every trial contributes an equal-length chunk to both series, the final
# concatenated clueGiver/guesser series are guaranteed equal length overall.
# Which words ended up in the trial set is reported rather than chosen by hand.
# condition_val: "board", "ground", or NULL to allow either
get_aligned_pair_n_trials <- function(data, dyad_val, metric, n_trials = 2, condition_val = NULL) {
  trial_pool <- data %>% filter(dyad == dyad_val)
  if (!is.null(condition_val)) trial_pool <- filter(trial_pool, condition == condition_val)

  trial_words <- trial_pool %>%
    distinct(word, session_datetime) %>%
    arrange(session_datetime) %>%
    slice_head(n = n_trials) %>%
    pull(word)

  cg_chunks <- vector("list", length(trial_words))
  gs_chunks <- vector("list", length(trial_words))

  for (i in seq_along(trial_words)) {
    pair <- get_aligned_pair(data, dyad_val, trial_words[i], metric)
    cg_chunks[[i]] <- pair$clueGiver
    gs_chunks[[i]] <- pair$guesser
  }

  cat("dyad:", dyad_val, "-", length(trial_words), "trial(s) used (chronological order):",
      paste(trial_words, collapse = ", "), "\n")

  list(
    clueGiver = unlist(cg_chunks, use.names = FALSE),
    guesser   = unlist(gs_chunks, use.names = FALSE),
    trials    = trial_words
  )
}
```

## Checking data alignment

```{r wms-sanity-check}
# checks that the trial resolves to exactly one dyad/word/condition, that
# both roles are present, and reports how far apart the two cams' start/end
# timestamps are (should be ~0, since both start recording together)
sanity_check_pair <- function(data, dyad_val, word_val) {
  trial <- data %>% filter(dyad == dyad_val, word == word_val)

  stopifnot(
    "trial resolves to more than one dyad" = n_distinct(trial$dyad) == 1,
    "trial resolves to more than one word" = n_distinct(trial$word) == 1,
    "trial resolves to more than one condition" = n_distinct(trial$condition) == 1,
    "both clueGiver and guesser must be present" = setequal(unique(trial$role), c("clueGiver", "guesser"))
  )

  cg <- trial %>% filter(role == "clueGiver") %>% arrange(frame)
  gs <- trial %>% filter(role == "guesser")   %>% arrange(frame)

  cat("dyad:", unique(trial$dyad), " word:", unique(trial$word), " condition:", unique(trial$condition), "\n")
  cat("frames  - clueGiver:", nrow(cg), " guesser:", nrow(gs), "\n")
  cat("start t - clueGiver:", cg$timestamp[1], "s  guesser:", gs$timestamp[1], "s",
      " (diff:", abs(cg$timestamp[1] - gs$timestamp[1]), "s)\n")
  cat("end t   - clueGiver:", tail(cg$timestamp, 1), "s  guesser:", tail(gs$timestamp, 1), "s",
      " (diff:", abs(tail(cg$timestamp, 1) - tail(gs$timestamp, 1)), "s)\n")

  invisible(list(clueGiver = cg, guesser = gs))
}
```

## WMS analysis - single trial

Note that here you can change the WMS parameters. 

```{r wms-run}
# single trial: one word, one dyad
trial_dyad <- "103_203"
trial_word <- "pasta_ground"
#doughnut_board
#pasta_ground

sanity_check_pair(df, trial_dyad, trial_word)

pair_x <- get_aligned_pair(df, trial_dyad, trial_word, "delta_gaze_angle_x")
pair_y <- get_aligned_pair(df, trial_dyad, trial_word, "delta_gaze_angle_y")

wms_x <- win_ms_synch(pair_x$clueGiver, pair_x$guesser,
                       srate = srate, win = 30, minfreq = 0.25, maxfreq = 10,
                       mincycle = 5, maxcycle = 8, nfreqs = 32, wavelet_length = 4)
wms_y <- win_ms_synch(pair_y$clueGiver, pair_y$guesser,
                       srate = srate, win = 30, minfreq = 0.25, maxfreq = 10,
                       mincycle = 5, maxcycle = 8, nfreqs = 32, wavelet_length = 4)
```

## WMS plots - single trial

```{r wms-plot}
plot_wms(wms_x, title_str = paste(trial_dyad, trial_word, "gaze x"),show_arrows  = TRUE,
         arrow_n_time = 20,
         arrow_n_freq = 6,
         arrow_scale  = 0.35,
         arrow_min_si = 0.4)
plot_wms(wms_y, title_str = paste(trial_dyad, trial_word, "gaze y"),show_arrows  = TRUE,
         arrow_n_time = 20,
         arrow_n_freq = 6,
         arrow_scale  = 0.35,
         arrow_min_si = 0.4)
```

## WMS analysis - multi-trial

Note that here you can change the WMS parameters. 


```{r wms-run-multi-trial}
# multiple trials, one dyad: concatenates the first n_trials words
# chronologically for trial_dyad (see get_aligned_pair_n_trials in wms-setup)
trial_dyad      <- "103_203"
n_trials        <- 5
trial_condition <- "board"  # "board", "ground", or NULL for either

pair_x <- get_aligned_pair_n_trials(df, trial_dyad, "delta_gaze_angle_x", n_trials = n_trials, condition_val = trial_condition)
pair_y <- get_aligned_pair_n_trials(df, trial_dyad, "delta_gaze_angle_y", n_trials = n_trials, condition_val = trial_condition)

nfreqs <- floor((log2(length(pair_x$clueGiver)) - 1) * 12)


wms_x_multi <- win_ms_synch(pair_x$clueGiver, pair_x$guesser,
                             srate = srate, win = 30, minfreq = 0.25, maxfreq = 10,
                             mincycle =5, maxcycle = 8, nfreqs = nfreqs, wavelet_length = 4)
wms_y_multi <- win_ms_synch(pair_y$clueGiver, pair_y$guesser,
                             srate = srate, win = 30, minfreq = 0.25, maxfreq = 10,
                             mincycle = 5, maxcycle = 8, nfreqs = nfreqs, wavelet_length = 4)
```

## WMS plots - multi-trial

```{r wms-plot-multi-trial}
trials_label <- paste(pair_x$trials, collapse = ", ")

plot_wms(wms_x_multi, title_str = paste0("On board - unaffiliated", " Δ gaze angle x trials n: (", n_trials, ")"),
         show_arrows  = TRUE,
         arrow_n_time = 20,
         arrow_n_freq = 6,
         arrow_scale  = 0.5,
         arrow_min_si = 0.4)
plot_wms(wms_y_multi, title_str = paste0("On board - unaffiliated", " Δ gaze angle y trials n: (", n_trials, ")"),
         show_arrows  = TRUE,
         arrow_n_time = 20,
         arrow_n_freq = 6,
         arrow_scale  = 0.5,
         arrow_min_si = 0.4)
```


# Creating matching WMS plots for mdRQA plots

## Unaffiliated couple, body movement WMS plots

```{r tkara data}
guesser <- read.csv("103_203_20_1_20250113_152735_bitter_board_guesser_cam01_body_displacement.csv")
clueGiver <- read.csv("103_203_20_1_20250113_152735_bitter_board_clueGiver_cam01_body_displacement.csv")
```

```{r}
nfreqs <- floor((log2(607) - 1) * 12)


wms_body <- win_ms_synch(guesser$displacement_smooth[1:607], clueGiver$displacement_smooth[1:607],
                       srate = srate, win = 30, minfreq = 0.25, maxfreq = 10,
                       mincycle = 5, maxcycle = 8, nfreqs = nfreqs, wavelet_length = 4 )


plot_wms(wms_body, title_str = "body displacement - bitter board,103-203",
        show_arrows  = TRUE,
         arrow_n_time = 20,
         arrow_n_freq = 6,
         arrow_scale  = 0.35,
         arrow_min_si = 0.4)

```

## Affiliated couple, body movement WMS plots


```{r tkara data}
guesser <- read.csv("116_216_45_1_20250116_103813_gravy_board_guesser_cam01_body_displacement.csv")
clueGiver <- read.csv("116_216_45_1_20250116_103813_gravy_board_clueGiver_cam01_body_displacement.csv")
```

```{r}
nfreqs <- floor((log2(607) - 1) * 12)


wms_body <- win_ms_synch(guesser$displacement_smooth[1:607], clueGiver$displacement_smooth[1:607],
                       srate = srate, win = 30, minfreq = 0.25, maxfreq = 10,
                       mincycle = 5, maxcycle = 8, nfreqs = nfreqs, wavelet_length = )


plot_wms(wms_body, title_str = "body displacement - bitter board,103-203",
        show_arrows  = TRUE,
         arrow_n_time = 20,
         arrow_n_freq = 6,
         arrow_scale  = 0.35,
         arrow_min_si = 0.4)

```

```{r}
# Mean SI within a frequency band
band  <- which(wms_body$freqs2use >= 0.25 & wms_body$freqs2use <= 5)

# Time course at a specific frequency
hz05_idx <- which.min(abs(wms_body$freqs2use - 5))
si_tc    <- wms_body$synch[hz05_idx, ]

# Peak synchrony location
pk <- which(wms_body$synch == max(wms_body$synch), arr.ind = TRUE)
```

```{r}
data.frame(
  Extraction = c("Band (0.25–5 Hz)", "Frequency (~5 Hz)", "Peak"),
  Metric     = c("Mean SI", "SD of SI over time at (~5 Hz)", "Peak SI"),
  Value      = c(round(mean(wms_body$synch[band, ]), 3),
                 round(sd(si_tc), 3),
                 round(max(wms_body$synch), 3)),
  Location   = c("—", "—",
                 paste0(round(wms_body$freqs2use[pk[1]], 2), " Hz @ ",
                        round(wms_body$time_synch[pk[2]], 1), " s"))
) |> knitr::kable(align = "llrl")
```


