---
title: "Categorical Cross-Recurrence of Gesture Classes (Guesser vs. Clue-Giver)"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

### Notes

- This is **cross-RQA**: recurrence = the two participants showing the *same* gesture
  class at (possibly different) time points. To instead run **auto-RQA** on a single
  participant (when that person repeats their own gesture class), pass the same series
  to both `ts1` and `ts2` (e.g. `ts1 = ts_guesser, ts2 = ts_guesser`) and set `tw = 1`
  to exclude the main diagonal.


```{r setup, include=FALSE}
# Global chunk options. echo = TRUE so the code is visible in the knitted output.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Load packages

```{r packages}
# readr / dplyr for fast, clean data wrangling; crqa for (categorical) RQA.
# Install once if needed:
#   install.packages(c("readr", "dplyr", "crqa"))
library(readr)
library(dplyr)
library(crqa)
```

## 2. Point to the two files

```{r paths}
# File1 = guesser, File2 = clue-giver. These are hard paths, they point to where they live on my machine lcoally. You need to copy the file path to the files on your computer and replace these. 

file_guesser   <- "/Users/tmullins/Documents/GitHub/TilburgMultiscaleSummerschool2026/Datasets/BalanceCorpus/gestureclassifications/preparation_103_203_5_guesser_cam02.mp4_predictions.csv"

file_clueGiver <- "/Users/tmullins/Documents/GitHub/TilburgMultiscaleSummerschool2026/Datasets/BalanceCorpus/gestureclassifications/preparation_103_203_5_clueGiver_cam02.mp4_predictions.csv"
```

## 3. Read in both files

```{r read}
# Read the full CSVs. We only need two columns from each, selected in the next step.
guesser_raw   <- read_csv(file_guesser)
clueGiver_raw <- read_csv(file_clueGiver)

# Quick sanity check on the columns available in each file.
names(guesser_raw)
names(clueGiver_raw)
```

## 4. Keep only `time` and `cnn_class`, and rename per participant

```{r select}
# Keep the two columns of interest and give cnn_class a participant-specific name
# so the two series don't collide when we merge.

guesser <- guesser_raw %>%
  select(time, cnn_class) %>%
  rename(cnn_class_guesser = cnn_class)

clueGiver <- clueGiver_raw %>%
  select(time, cnn_class) %>%
  rename(cnn_class_clueGiver = cnn_class)
```

## 5. Merge on matched time

```{r merge}
# Inner join keeps only rows where both participants share the same `time`,
# so every row is a matched time point with one cnn_class column per participant.

merged <- inner_join(guesser, clueGiver, by = "time") %>%
  arrange(time)

# If the two files were sampled at the same frame rate, times should line up
# exactly. If tiny floating-point differences drop rows, round first, e.g.:
#   mutate(time = round(time, 3)) on both frames before joining.

# Report how many time points were matched.
cat("Matched time points:", nrow(merged), "\n")
head(merged)
```

## 6. Encode the categorical classes as integers (shared coding scheme)

```{r encode}
# crqa needs numeric input. Build ONE factor level set from the union of classes
# across both participants so an identical class maps to the same integer code in
# both series (essential for cross-recurrence to be meaningful).

all_levels <- union(merged$cnn_class_guesser, merged$cnn_class_clueGiver)

merged <- merged %>%
  mutate(
    code_guesser   = as.integer(factor(cnn_class_guesser,   levels = all_levels)),
    code_clueGiver = as.integer(factor(cnn_class_clueGiver, levels = all_levels))
  )

# The two numeric series that go into cRQA.
ts_guesser   <- merged$code_guesser
ts_clueGiver <- merged$code_clueGiver

# Mapping of integer code -> class label, for reference.
data.frame(code = seq_along(all_levels), cnn_class = all_levels)
```

## 7. Run categorical cross-RQA

```{r crqa}
# Categorical RQA settings:
#   delay = 1, embed = 1  -> no time-delay embedding (correct for nominal data)
#   rescale = 0, normalize = 0 -> do not rescale because we use integer codes
#   radius = 0.0001 -> only exact code matches count as recurrences (1 is different from 2)
#                      (distinct integer codes differ by >= 1)
#   tw = 0 -> Theiler window off; keep the diagonal (meaningful in cross-recurrence)
#   datatype = "categorical"

res <- crqa(
  ts1        = ts_guesser,
  ts2        = ts_clueGiver,
  delay      = 1,
  embed      = 1,
  rescale    = 0,
  radius     = 0.0001,
  normalize  = 0,
  mindiagline = 2,
  minvertline = 2,
  tw         = 0,
  whiteline  = FALSE,
  recpt      = FALSE,
  side       = "both",
  method     = "crqa",
  metric     = "euclidean",
  datatype   = "categorical"
)

# RQA measures: RR (recurrence rate), DET (determinism), maxL/L, ENTR, LAM, TT, etc.
res[!names(res) %in% "RP"]
```

## 8. Plot the recurrence plot (RP)

```{r}
library(ggplot2)
library(reshape2)

create_cross_rp_plot <- function(rp, graph_title,
                                 xlab = "Guesser (time point)",
                                 ylab = "Clue-giver (time point)") {
  # Convert RP matrix to long format
  rqpa.data <- reshape2::melt(as.matrix(rp))
  
  # Filter for recurrence points
  rqpa.data <- rqpa.data[rqpa.data$value == TRUE | rqpa.data$value == 1, ]
  
  ggplot(data = rqpa.data, aes(x = Var1, y = Var2)) +
    geom_point(size = 0.5, color = 'black') +   # Small black points for recurrences
    labs(x = xlab, y = ylab, title = graph_title) +
    theme_minimal() +
    theme(axis.text.x  = element_text(size = 9, angle = 0, vjust = 0.3),
          axis.text.y  = element_text(size = 9),
          axis.text    = element_text(size = 16),
          axis.title   = element_text(size = 16),
          plot.title   = element_text(size = 16, hjust = 0.5),
          aspect.ratio = 1,
          panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6)) +
    coord_fixed()
}

p <- create_cross_rp_plot(res$RP, "Gesture Class Cross-Recurrence\nPreparation 103_203_5")
print(p)

# Optional: save
# ggsave("CrossRP_103_203_5.png", plot = p, width = 6, height = 6, dpi = 300)
```
---

```{r}
# --- mdRQA: dyad as ONE system -------------------------------------------
# Unlike the cross-RQA above (which only fires when the two people share a class),
# mdRQA treats (guesser, clueGiver) as a single 2-D state. State 12 at time i
# recurring as 12 at time j counts, even though they're doing different things.
# So all joint configurations -- 11, 12, 21, 22, ... -- are states in their own right.

## Encoding A: integer codes -> EXACT joint-state recurrence
md_int <- as.matrix(merged[, c("code_guesser", "code_clueGiver")])

## Encoding B: one-hot -> GRADED similarity (see radius note below)
onehot <- function(x, levs) {
  m <- matrix(0L, nrow = length(x), ncol = length(levs))
  m[cbind(seq_along(x), match(x, levs))] <- 1L
  m
}
md_oh <- cbind(
  onehot(merged$cnn_class_guesser,   all_levels),
  onehot(merged$cnn_class_clueGiver, all_levels)
)

# Under one-hot + Euclidean, joint-state distances are discrete and interpretable:
#   both participants match  -> 0
#   exactly one mismatches   -> sqrt(2) ~= 1.414   <- "similar" state
#   both mismatch            -> 2
# So:
#   radius = 0.0001 -> exact joint state only (equivalent to Encoding A)
#   radius = 1.5    -> exact PLUS "one of the two matches"  <- what you want
radius_similar <- 1.5

# tw = 1: this is AUTO-recurrence of one system, so exclude the line of identity
res_md <- crqa(
  ts1         = md_oh,
  ts2         = md_oh,
  delay       = 1,      # no embedding: the dimensions ARE the system
  embed       = 1,
  rescale     = 0,
  radius      = radius_similar,
  normalize   = 0,
  mindiagline = 2,
  minvertline = 2,
  tw          = 1,
  whiteline   = FALSE,
  recpt       = FALSE,
  side        = "both",
  method      = "mdcrqa",
  metric      = "euclidean"
)

res_md[!names(res_md) %in% "RP"]

# Reuse the plot function from the previous chunk
p_md <- create_cross_rp_plot(
  res_md$RP,
  "Dyadic mdRQA: revisited joint gesture states",
  xlab = "Time point",
  ylab = "Time point"
)
print(p_md)

```

```{r}
# How often does each joint state occur? Sanity-check before trusting RR/DET.
merged %>%
  count(cnn_class_guesser, cnn_class_clueGiver, sort = TRUE) %>%
  mutate(prop = round(n / sum(n), 3))

# Radius sweep: RR should jump at 0, sqrt(2), and 2 -- nowhere else.
sapply(c(0.0001, 1.5, 2.5), function(r) {
  crqa(md_oh, md_oh, delay = 1, embed = 1, rescale = 0, radius = r,
       normalize = 0, mindiagline = 2, minvertline = 2, tw = 1,
       method = "mdcrqa", metric = "euclidean")$RR
})
```

```{r}
library(dplyr)
library(ggplot2)
library(reshape2)

# Label each time point with its joint state, e.g. "point|beat"
merged <- merged %>%
  mutate(joint_state = paste(cnn_class_guesser, cnn_class_clueGiver, sep = "|"))

state_lookup <- merged %>%
  mutate(ID = row_number()) %>%
  select(ID, joint_state, cnn_class_guesser, cnn_class_clueGiver)

# Melt the RP and keep only recurrent points
rp_long <- reshape2::melt(as.matrix(res_md$RP))
rp_long <- rp_long[rp_long$value == TRUE | rp_long$value == 1, ]

# Attach the state at BOTH ends of each recurrence
rp_long <- rp_long %>%
  left_join(state_lookup, by = c("Var1" = "ID")) %>%
  rename(state_i = joint_state, g_i = cnn_class_guesser, c_i = cnn_class_clueGiver) %>%
  left_join(state_lookup, by = c("Var2" = "ID")) %>%
  rename(state_j = joint_state, g_j = cnn_class_guesser, c_j = cnn_class_clueGiver) %>%
  mutate(
    # Because radius = 1.5, a recurrence is either an exact joint-state revisit
    # or a partial one (exactly one participant matches).
    match_type = case_when(
      state_i == state_j                 ~ "Exact joint state",
      g_i == g_j & c_i != c_j            ~ "Guesser matches only",
      c_i == c_j & g_i != g_j            ~ "Clue-giver matches only",
      TRUE                               ~ "Other"
    )
  )

# --- Option 1: colour by match type (unambiguous under radius = 1.5) -------
p_md_type <- ggplot(rp_long, aes(x = Var1, y = Var2, fill = match_type)) +
  geom_raster() +
  scale_fill_manual(values = c(
    "Exact joint state"       = "#000000",
    "Guesser matches only"    = "#D55E00",
    "Clue-giver matches only" = "#0072B2",
    "Other"                   = "grey80"
  )) +
  labs(x = "Time point", y = "Time point",
       title = "Dyadic mdRQA: revisited joint states",
       fill = "Recurrence type") +
  theme_minimal() +
  theme(axis.text.x  = element_text(size = 9, angle = 0, vjust = 0.3),
        axis.text.y  = element_text(size = 9),
        axis.title   = element_text(size = 16),
        plot.title   = element_text(size = 16, hjust = 0.5),
        legend.title = element_text(size = 10),
        aspect.ratio = 1,
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6)) +
  coord_fixed()

print(p_md_type)

# --- Option 2: colour by the joint state itself ---------------------------
# Only exact recurrences get a state colour; partial ones stay grey, since
# their two endpoints disagree and no single state colour would be honest.
p_md_state <- rp_long %>%
  mutate(state_col = ifelse(state_i == state_j, state_i, NA)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = state_col)) +
  geom_raster() +
  scale_fill_discrete(na.value = "grey85") +
  labs(x = "Time point", y = "Time point",
       title = "Dyadic mdRQA: exact joint-state revisits",
       fill = "Joint state\n(guesser|clue-giver)") +
  theme_minimal() +
  theme(axis.title   = element_text(size = 16),
        plot.title   = element_text(size = 16, hjust = 0.5),
        legend.title = element_text(size = 10),
        legend.text  = element_text(size = 8),
        aspect.ratio = 1,
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6)) +
  coord_fixed()

print(p_md_state)
```

```{r}
library(dplyr)

# Option 2 greys a cell whenever state_i != state_j (that's the ifelse -> NA).
# So: every recurrent pair that is NOT an exact joint-state revisit.
grey_cells <- rp_long %>%
  filter(state_i != state_j)

cat("Recurrent cells total:", nrow(rp_long), "\n")
cat("Grey (non-exact) cells:", nrow(grey_cells),
    sprintf("(%.1f%%)\n", 100 * nrow(grey_cells) / nrow(rp_long)))

# --- Which state PAIRS are being greyed, and how often ---------------------
# Unordered: (A,B) and (B,A) are the same pairing, so collapse them.
grey_pairs <- grey_cells %>%
  mutate(
    s1 = pmin(state_i, state_j),
    s2 = pmax(state_i, state_j)
  ) %>%
  count(s1, s2, match_type, sort = TRUE) %>%
  mutate(prop_of_grey = round(n / sum(n), 3))

grey_pairs

# --- Same thing, but broken out so you can see WHO matched -----------------
grey_detail <- grey_cells %>%
  count(state_i, state_j, match_type, sort = TRUE) %>%
  mutate(prop = round(n / sum(n), 3))

head(grey_detail, 20)

# --- Sanity check: does the split match the plot? --------------------------
# Exact = coloured in Option 2; everything else = grey.
rp_long %>%
  mutate(coloured = ifelse(state_i == state_j, "Coloured (exact)", "Grey (partial)")) %>%
  count(coloured) %>%
  mutate(prop = round(n / sum(n), 3))
```