---
title: "mdrqa_displacement"
author: "TkaraMullins"
date: "2026-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load packages

```{r packages}
# readr / dplyr for fast, clean data wrangling; crqa for RQA.
# Install once if needed:
#   install.packages(c("readr", "dplyr", "crqa", "ggplot2"))
library(readr)
library(dplyr)
library(crqa)
library(ggplot2)
```

## 2. Point to the files

```{r paths, include=FALSE}
#guesser_body   <- read_csv("/Users/tmullins/Desktop/motiontest/103_203_20_1_20250113_152735_bitter_board_guesser_cam01_body_displacement.csv")
#clueGiver_body <- read_csv("/Users/tmullins/Desktop/motiontest/103_203_20_1_20250113_152735_bitter_board_clueGiver_cam01_body_displacement.csv")
#gaze_ground_all <- read_csv("/Users/tmullins/Downloads/gaze_angle_ground.csv")
#gaze_board_all  <- read_csv("/Users/tmullins/Downloads/gaze_angle_board.csv")
```

```{r paths, include=FALSE}
guesser_body   <- read_csv("/Users/tmullins/Desktop/motiontest/116_216_45_1_20250116_103813_gravy_board_guesser_cam01_body_displacement.csv")
clueGiver_body <- read_csv("/Users/tmullins/Desktop/motiontest/116_216_45_1_20250116_103813_gravy_board_clueGiver_cam01_body_displacement.csv")
#gaze_ground_all <- read_csv("/Users/tmullins/Downloads/gaze_angle_ground.csv")
#gaze_board_all  <- read_csv("/Users/tmullins/Downloads/gaze_angle_board.csv")
```

```{r names, eval=FALSE}
names(guesser_body)
names(clueGiver_body)
names(gaze_ground_all)
names(gaze_board_all)
```

Convert body time from milliseconds to seconds so it's on the same unit as gaze.

```{r body-time}
clueGiver_body <- clueGiver_body %>%
  mutate(time = round(time / 1000, 3))

guesser_body <- guesser_body %>%
  mutate(time = round(time / 1000, 3))
```

## 3. Filter for the proper trial

Keep `role`, `timestamp`, `delta_gaze_angle_x` and `delta_gaze_angle_y` for mdRQA.
Rename `timestamp` to `time` to match the displacement files.

```{r select-gaze}
TRIAL <- "bitter_board"   # single place to change the trial

gaze_board_trial_clueGiver <- gaze_board_all %>%
  filter(word == TRIAL, role == "clueGiver") %>%
  select(role, timestamp, delta_gaze_angle_x, delta_gaze_angle_y) %>%
  rename(time = timestamp)

gaze_board_trial_guesser <- gaze_board_all %>%
  filter(word == TRIAL, role == "guesser") %>%
  select(role, timestamp, delta_gaze_angle_x, delta_gaze_angle_y) %>%
  rename(time = timestamp)
```

## 4. Collapse gaze x and y into one stream per person

```{r gaze-composite}
gaze_board_trial_clueGiver <- gaze_board_trial_clueGiver %>%
  mutate(clueGiver_gaze = sqrt(delta_gaze_angle_x^2 + delta_gaze_angle_y^2))

gaze_board_trial_guesser <- gaze_board_trial_guesser %>%
  mutate(guesser_gaze = sqrt(delta_gaze_angle_x^2 + delta_gaze_angle_y^2))
```

Drop `role` — it's redundant once it's in the column name, and it would collide as
`role.x` / `role.y` in the joins below.

```{r trim-gaze}
gaze_board_trial_clueGiver <- gaze_board_trial_clueGiver %>%
  select(time, clueGiver_gaze)

gaze_board_trial_guesser <- gaze_board_trial_guesser %>%
  select(time, guesser_gaze)
```

Same idea for body: rename the displacement column per person BEFORE joining,
otherwise the merged frame gets `displacement.x` / `displacement.y` with no way to
tell them apart.

```{r trim-body}
DISP_COL <- "displacement"   # <- edit to the real column name

clueGiver_body_trim <- clueGiver_body %>%
  select(time, clueGiver_body = all_of(DISP_COL))

guesser_body_trim <- guesser_body %>%
  select(time, guesser_body = all_of(DISP_COL))
```

## 5. Merge on matched time

Inner join keeps only rows where both participants share the same `time`, so every
row is a matched time point with one column per stream.

```{r merge-gaze}
merged <- inner_join(gaze_board_trial_guesser, gaze_board_trial_clueGiver, by = "time") %>%
  arrange(time)

cat("Gaze matched time points:", nrow(merged), "\n")
```

```{r merge-body}
mdrqa_df <- merged %>%
  inner_join(guesser_body_trim,   by = "time") %>%
  inner_join(clueGiver_body_trim, by = "time") %>%
  arrange(time) %>%
  select(time, clueGiver_body, clueGiver_gaze, guesser_body, guesser_gaze)

cat("Final matched time points:", nrow(mdrqa_df), "\n")
head(mdrqa_df)
```

## 6. Build the multivariate matrix

mdRQA treats the dyad as one system: each column is a dimension, each row a time
point.

`DIMS` is the one place that decides what goes into the state space. Your earlier
version described four dimensions in prose but selected only the two body columns —
this makes the choice explicit so the two can't drift apart again. Uncomment the
four-column version if gaze belongs in the analysis.

```{r md-matrix}
DIMS <- c("clueGiver_body", "guesser_body")
# DIMS <- c("clueGiver_body", "clueGiver_gaze", "guesser_body", "guesser_gaze")

M <- mdrqa_df %>%
  select(all_of(DIMS)) %>%
  as.matrix()

# Body displacement and gaze angle live on different scales and units, so z-score
# each column. Without that, whichever stream has the larger variance dominates the
# Euclidean distance and the others contribute almost nothing.
M <- scale(M)

FS <- 30        # sampling rate in Hz, from the 0.033 s spacing
cat("Matrix:", nrow(M), "samples x", ncol(M), "dimensions (~",
    round(nrow(M) / FS, 1), "s )\n")
```

## 7. Set parameters

No embedding: the simultaneous streams are taken as the state space directly, so
`embed = 1` and delay is inert.

```{r params}
DELAY <- 1
EMBED <- 1

run_md <- function(radius) {
  crqa(M, M,
       delay = DELAY, embed = EMBED, rescale = 1, radius = radius,
       normalize = 0, mindiagline = 2, minvertline = 2, tw = 0,
       whiteline = FALSE, recpt = FALSE, side = "both",
       method = "mdcrqa", metric = "euclidean", datatype = "continuous")
}
```

## 8. Sweep radius to hit a fixed RR

RR is held constant so the line-based measures are comparable across trials. Radius
is whatever it takes to get there, and gets reported as a nuisance parameter rather
than interpreted.

```{r radius-sweep}
target_rr <- 5   # hold this identical for every trial

rr_at <- function(r) tryCatch(run_md(r)$RR, error = function(e) NA_real_)

# Coarse pass to bracket the target
coarse <- seq(0.05, 3, by = 0.05)
rrs_c  <- sapply(coarse, rr_at)

plot(coarse, rrs_c, type = "b", xlab = "radius", ylab = "RR (%)")
abline(h = target_rr, lty = 2)

# Fine pass around the bracket
i   <- which.min(abs(rrs_c - target_rr))
lo  <- coarse[max(i - 1, 1)]
hi  <- coarse[min(i + 1, length(coarse))]

fine  <- seq(lo, hi, length.out = 60)
rrs_f <- sapply(fine, rr_at)

best   <- which.min(abs(rrs_f - target_rr))
RADIUS <- fine[best]

cat("radius =", round(RADIUS, 4), "-> RR =", round(rrs_f[best], 3), "%\n")
```

```{r trial-level}
res <- run_md(RADIUS)

mdrqa_out <- data.frame(
  trial  = TRIAL,
  n      = nrow(M),
  radius = RADIUS,
  RR     = res$RR,    # fixed at target — confirms the sweep landed
  DET    = res$DET,
  LAM    = res$LAM,
  ENTR   = res$ENTR
)

print(mdrqa_out)
```

```{r recurrence-plot}
rp_df <- as.data.frame(which(as.matrix(res$RP), arr.ind = TRUE))
names(rp_df) <- c("x", "y")
rp_df$x <- rp_df$x / FS
rp_df$y <- rp_df$y / FS

ggplot(rp_df, aes(x = x, y = y)) +
  geom_point(shape = 15, size = 0.3, colour = "black") +
  coord_fixed(xlim = c(0, nrow(M) / FS), ylim = c(0, nrow(M) / FS), expand = FALSE) +
  labs(x = "Time (s)", y = "Time (s)", title = "mdRQA recurrence plot") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, colour = "grey70"))
```

## 9. Sliding-window mdRQA

Windows are analysed at a *fixed RR*, not a fixed radius. With a fixed radius the
recurrence density drifts from window to window and DET rides along with it, so
"elevated DET" would really mean "elevated RR". Holding RR constant makes DET a
statement about the *structure* of recurrence — how much of it falls on diagonal
lines — rather than the amount.

Because RR is just the proportion of pairwise distances at or below the radius, the
radius that yields `TARGET_RR` is the `TARGET_RR`-th percentile of that window's own
distance matrix. That's exact and costs one `crqa` call per window instead of a
sweep.

One bookkeeping note: these windows use `rescale = 0` and a raw-distance radius,
while section 8 uses `rescale = 1` (radius in units of mean distance). The two
`radius` numbers therefore aren't on the same scale and shouldn't be compared to
each other. RR — the thing actually held fixed — is comparable throughout.

```{r}
WIN_SEC   <- 2      # window length
STEP_SEC  <- 1   # hop — keep it below the window
TARGET_RR <- 5
THEILER   <- 1      # exclude the line of identity

WIN  <- round(WIN_SEC  * FS)
STEP <- round(STEP_SEC * FS)

stopifnot(nrow(M) >= WIN)
starts <- seq(1, nrow(M) - WIN + 1, by = STEP)
cat(length(starts), "windows of", WIN, "samples, hop", STEP, "samples\n")
cat("LOI would be", round(100 / WIN, 2), "% RR if not excluded\n")
```

```{r window-fn}
# One window -> mdRQA at fixed RR. Returns an NA row on failure rather than
# aborting the whole loop.
win_md <- function(s) {
  idx <- s:(s + WIN - 1)
  W   <- M[idx, , drop = FALSE]

  # Radius = quantile of this window's distance matrix.
  # Full N x N including the zero diagonal, matching crqa's RR with tw = 0.
  D <- as.matrix(dist(W, method = "euclidean"))
  r <- as.numeric(quantile(D, probs = TARGET_RR / 100, names = FALSE))

  # Degenerate window: more than TARGET_RR% of distances are exactly 0 (a frozen
  # signal, or dropped frames carried forward). No radius hits the target RR.
  if (!is.finite(r) || r <= 0) {
    return(data.frame(start = s, radius = NA_real_, RR = NA_real_,
                      DET = NA_real_, LAM = NA_real_, ENTR = NA_real_,
                      flat = TRUE))
  }

  res_w <- tryCatch(
    crqa(W, W,
         delay = DELAY, embed = EMBED, rescale = 0, radius = r,
         normalize = 0, mindiagline = 2, minvertline = 2, tw = THEILER,
         whiteline = FALSE, recpt = FALSE, side = "both",
         method = "mdcrqa", metric = "euclidean", datatype = "continuous"),
    error = function(e) NULL
  )
  if (is.null(res_w)) {
    return(data.frame(start = s, radius = r, RR = NA_real_, DET = NA_real_,
                      LAM = NA_real_, ENTR = NA_real_, flat = FALSE))
  }

  data.frame(start = s, radius = r, RR = res_w$RR, DET = res_w$DET,
             LAM = res_w$LAM, ENTR = res_w$ENTR, flat = FALSE)
}
```

```{r window-run}
win_df <- do.call(rbind, lapply(starts, win_md))

# Map sample indices back to real timestamps from mdrqa_df.
win_df <- win_df %>%
  mutate(
    end        = start + WIN - 1,
    start_time = mdrqa_df$time[start],
    end_time   = mdrqa_df$time[end],
    mid_time   = (start_time + end_time) / 2
  )

# Sanity check: the quantile trick should land RR on target in every window.
cat("RR range:", round(range(win_df$RR, na.rm = TRUE), 3), "\n")
cat("Windows dropped:", sum(is.na(win_df$DET)),
    "( flat:", sum(win_df$flat), ")\n")
```

## 10. Flag windows with elevated DET

The threshold is the within-trial mean + 2 SD of the windowed DET distribution. Two
caveats worth keeping in view:

- **This is relative, not absolute.** Every trial flags *something*, because the
  threshold is built from that trial's own mean. A flat, uneventful trial still
  produces "peaks". If the claim is that these moments are meaningful rather than
  merely above-average, the threshold needs a surrogate baseline — shuffle one
  participant's window order, or pair non-interacting participants, and rebuild the
  cutoff from that null.
- **DET is usually left-skewed and bounded at 100.** Mean + 2 SD assumes rough
  symmetry. Check the histogram; if it's badly skewed, the median + 2·MAD version is
  the more honest cutoff. Both are computed — pick one and say which in the write-up.

```{r threshold}
det <- win_df$DET

mu  <- mean(det, na.rm = TRUE)
sdv <- sd(det,   na.rm = TRUE)
thresh_sd <- mu + 2 * sdv

med        <- median(det, na.rm = TRUE)
madv       <- mad(det, na.rm = TRUE)
thresh_mad <- med + 2 * madv

hist(det, breaks = 30, main = "Windowed DET", xlab = "DET (%)")
abline(v = thresh_sd,  col = "red",  lwd = 2)
abline(v = thresh_mad, col = "blue", lwd = 2, lty = 2)
legend("topleft", c("mean + 2SD", "median + 2MAD"),
       col = c("red", "blue"), lty = c(1, 2), bty = "n")

THRESH <- thresh_sd   # <- switch to thresh_mad if the histogram is skewed

win_df$elevated <- !is.na(det) & det > THRESH
cat("Threshold:", round(THRESH, 2), "% DET |",
    sum(win_df$elevated), "of", sum(!is.na(det)), "windows flagged\n")
```

```{r det-timecourse}
ggplot(win_df, aes(x = mid_time, y = DET)) +
  geom_line(colour = "grey40") +
  geom_point(data = subset(win_df, elevated), colour = "red", size = 1.2) +
  geom_hline(yintercept = THRESH, linetype = 2, colour = "red") +
  labs(x = "Time (s)", y = "DET (%)",
       title = paste0("Windowed DET (", WIN_SEC, "s window, ",
                      STEP_SEC, "s hop, RR fixed at ", TARGET_RR, "%)")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

## 11. Merge contiguous flags into events and save

Adjacent windows overlap heavily — a 5 s window with a 0.5 s hop shares 90% of its
samples with its neighbour — so a single burst of coordination shows up as a run of
consecutive flags. Collapse each run into one event spanning from the first sample
of its first window to the last sample of its last window.

```{r events}
rle_flag <- rle(win_df$elevated)
ends     <- cumsum(rle_flag$lengths)
begins   <- ends - rle_flag$lengths + 1
keep     <- which(rle_flag$values)

events <- data.frame(
  trial      = TRIAL,
  event      = seq_along(keep),
  start_time = win_df$start_time[begins[keep]],
  end_time   = win_df$end_time[ends[keep]],
  n_windows  = rle_flag$lengths[keep],
  peak_DET   = sapply(keep, function(k) max(win_df$DET[begins[k]:ends[k]])),
  mean_DET   = sapply(keep, function(k) mean(win_df$DET[begins[k]:ends[k]]))
) %>%
  mutate(duration = end_time - start_time)

print(events)

write_csv(win_df, paste0("mdrqa_windows_", TRIAL, ".csv"))   # every window
write_csv(events, paste0("mdrqa_events_",  TRIAL, ".csv"))   # flagged episodes only
```

Note what a timestamp means here: an event's `start_time` / `end_time` bound the
*windows* that were elevated, not a precise onset. Temporal resolution is the window
length — a 5 s window smears any instant of high determinism across roughly ±2.5 s.
Tighter localisation means shortening `WIN_SEC`, but not so far that each window
holds too few recurrence points for DET to be stable.
